Constructing Biological Regulatory Networks from Process Hitting models \\\rewriteil{Or possibly: “Complementarity between PH, IG and Thomas modeling”}

### 2013-07-04 parts/infer-ig.tex
fix mistakes in IG inference

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
The main idea of this inference is that
the presence of a positive (negative) influence of a component $b$ on $a$ denotes the fact that
there exists a state in which increasing the level of $b$ tends to make the future level of $a$ rise (drop)
(\pref{eq:edges-inference-pos} \& \eqref{eq:edges-inference-neg}).
Therefore, these influences are split into positive and negative ones, and represent possible edges in the final IG.
We ignore the cases where a component has no visible influence on another.
%
\begin{proposition}[Influences inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp. $\hat{E}_-$) for any $a\in\Gamma$ by:
\begin{align}
\begin{split}\label{eq:edges-inference-pos}
  \forall b\in\PHpredecgene{a}, \qquad\qquad\quad\\
  b\xrightarrow{t+1} a \in \hat{E}_+ \Longleftrightarrow\ &
    \exists \sigma \in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c, \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}),\\
    &\qquad \warnul{a_j \leq \PHget{\sigma}{a} < a_k \quad^*}\\
    &\qquad \warnul{a_j \leq \PHget{\sigma}{a} < a_k \vee a_j < \PHget{\sigma}{a} \leq a_k \quad^{**}}\\
    &\qquad \warnul{a_j < a_k \quad^{***}}
\end{split}\\
\begin{split}\label{eq:edges-inference-neg}
  \forall b\in\PHpredecgene{a}, \qquad\qquad\quad\\
  b\xrightarrow{t+1} a \in \hat{E}_- \Longleftrightarrow\ &
    \exists \sigma \in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c, \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}),\\
    &\qquad \warnul{a_j > \PHget{\sigma}{a} \geq a_k \quad^*}\\
    &\qquad \warnul{a_j > \PHget{\sigma}{a} \geq a_k \vee a_j \geq \PHget{\sigma}{a} > a_k \quad^{**}}\\
    &\qquad \warnul{a_j > a_k \quad^{***}}
\end{split}
\end{align}
\end{proposition}
%
\todo{\underline{$^*$Strict inference}}\\
\todo{\underline{$^{**}$Large inference}}\\
\todo{\underline{$^{***}$Non-unitary-style inference}}

@@

We point out that one of the benefits of this alternative inference is the fact that the conditions used to infer each influence
in \pref{eq:edges-inference-pos} \& \eqref{eq:edges-inference-neg}
allow to apply some results about the links between the structure of an IG and the dynamics of the model.
These results consist mainly in Thomas' conjectures, that focus on the presence of positive and negative circuits in the IG, and state that:
\begin{itemize}
  \item the presence of a positive circuit in the IG is a necessary condition for several fixed points,
  \item the presence of a negative circuit in the IG is a necessary condition for sustained oscillations.
\end{itemize}
The sign of a circuit is defined by the number of negative edges it contains;
given the nature of unsigned edges, the presence of such an edge in a circuit would count for the presence of both a positive and negative circuit.
These conjectures have been demonstrated in~\cite{Richard2010378} \towrite{Other references! (multi-stationarity \& boolean models)}.

>>>>

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
The main idea of this inference is that
the presence of a positive (negative) influence of a component $b$ on $a$ denotes the fact that
there exists a state in which increasing the level of $b$ tends to make the future level of $a$ rise (drop)
(\pref{eq:edges-inference}).
Therefore, these influences are split into positive and negative ones, and represent possible edges in the final IG.
Furthermore, studying the influences of the groups of regulators of $a$
allows to study its auto-influences, and thus infer auto-edges on $a$ in the IG (\pref{eq:edges-inference-auto}).
Finally, \pref{eq:edges-inference-noreg} handles the special case where $a$ has no regulators.
We ignore the cases where a component has no visible influence on another.
%
\begin{proposition}[Influences inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp. $\hat{E}_-$) for any $a\in\Gamma$ by:
\begin{align}
\begin{split}\label{eq:edges-inference}
  \forall b\in\PHpredecgene{a}, \forall s \in \{ +, - \}, \\
  b \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \exists g \in X(a), b \in g,
  \exists \sigma \in \textstyle\prod_{c\in g \cup \{ a \}} \PHl_c, \\
    &\qquad \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), \\
    &\qquad s = \f{sign}(a_k - a_j)
\end{split}\\
\begin{split}\label{eq:edges-inference-auto}
  \forall s \in \{ +, - \}, \quad\qquad\qquad \\
  a \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \exists g \in X(a), \exists b \in g,
  \exists \sigma \in \textstyle\prod_{c\in g \cup \{ a \}} \PHl_c, \\
    &\qquad \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), \\
    &\qquad s = \f{sign}(a_k - a_j)
\end{split}\\
\begin{split}\label{eq:edges-inference-noreg}
  \forall s \in \{ +, - \}, \quad\qquad\qquad \\
  a \xrightarrow{t+1} a \in \hat{E}_s \Longleftrightarrow\ & \reg{a} = \emptyset \wedge \{ a_t, a_{t+1} \} \subset \PHl_a, \\
    &\qquad \exists a_j \in \irB_a(\PHetat{a_t}), \exists a_k \in \irB_a(\PHetat{a_{t+1}}), \\
    &\qquad s = \f{sign}(a_k - a_j)
\end{split}
%\begin{split}\label{eq:edges-inference-pos}
%  \forall b\in\PHpredecgene{a}, \qquad\qquad\quad\\
%  b\xrightarrow{t+1} a \in \hat{E}_+ \Longleftrightarrow\ &
%    \exists \sigma \in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c, \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
%    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), a_j < a_k
%%    &\qquad \warnul{a_j \leq \PHget{\sigma}{a} < a_k \quad^*}\\
%%    &\qquad \warnul{a_j \leq \PHget{\sigma}{a} < a_k \vee a_j < \PHget{\sigma}{a} \leq a_k \quad^{**}}\\
%%    &\qquad \warnul{a_j < a_k \quad^{***}}
%\end{split}\\
%\begin{split}\label{eq:edges-inference-neg}
%  \forall b\in\PHpredecgene{a}, \qquad\qquad\quad\\
%  b\xrightarrow{t+1} a \in \hat{E}_- \Longleftrightarrow\ &
%    \exists \sigma \in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c, \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
%    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}), a_j > a_k
%%    &\qquad \warnul{a_j > \PHget{\sigma}{a} \geq a_k \quad^*}\\
%%    &\qquad \warnul{a_j > \PHget{\sigma}{a} \geq a_k \vee a_j \geq \PHget{\sigma}{a} > a_k \quad^{**}}\\
%    %&\qquad \warnul{a_j > a_k \quad^{***}}
%\end{split}
\end{align}
where $\f{sign}(x) = \begin{cases}\frac{|x|}{x} & \text{ if $x \neq 0$} \\ 0 & \text{ if $x = 0$}\end{cases}$.
\end{proposition}
%
%\todo{\underline{$^*$Strict inference}}\\
%\todo{\underline{$^{**}$Large inference}}\\
%\todo{\underline{$^{***}$Non-unitary-style inference}}

@@

We point out that one of the benefits of this inference is the fact that the conditions used to infer each influence
in \pref{pps:inference-edges}
allow to apply some results about the links between the structure of an IG and the dynamics of the model.
These results consist mainly in Thomas' conjectures, that focus on the presence of positive and negative circuits in the IG, and state that:
\begin{itemize}
  \item the presence of a positive circuit in the IG is a necessary condition for several fixed points,
  \item the presence of a negative circuit in the IG is a necessary condition for sustained oscillations.
\end{itemize}
The sign of a circuit is defined by the number of negative edges it contains;
given the nature of unsigned edges, the presence of such an edge in a circuit would count for the presence of both a positive and negative circuit.
These conjectures have been demonstrated in~\cite{Richard2010378} \towrite{Other references! (multi-stationarity \& boolean models)}.
